---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

```{r dataload, echo=TRUE}
# unzip the data file
unzip("activity.zip");

# read the file in
rawActivity<-read.csv("activity.csv");

# sanity check on the load
names(rawActivity);
dim(rawActivity);

# make a set with no NA's in the step column
validStepsOnly = rawActivity[!is.na(rawActivity$steps),];

```









## What is mean total number of steps taken per day?

First we will plot the total number of steps per day, excluding any NA values for steps.


```{r meanSteps, echo=TRUE}

library(plyr)


# summarise the steps for each day and add back to the frame
sumValidSteps<-ddply(validStepsOnly,c("date"),summarise, sumSteps=sum(steps))



# make a histogram of the step sums
hist(sumValidSteps$sumSteps, main="Histogram of Total Number of Steps Taken Each Day",
     xlab="Total Steps per Day");

```


```{r computeMeanAndMedian, echo=TRUE}
# compute the mean and median for reporting in the main text
totalStepsMean <- mean(sumValidSteps$sumSteps);
totalStepsMedian <- median(sumValidSteps$sumSteps);

```

The mean of the total number of steps per day is `r totalStepsMean`.

The median of the total number of steps per day is `r totalStepsMedian`





## What is the average daily activity pattern?

```{r averageSteps, echo=TRUE}

# get the mean steps per interval
averageStepsPerInterval <- ddply(validStepsOnly,c("interval"),summarise, meanSteps=mean(steps))

# plot it
plot(averageStepsPerInterval$interval, averageStepsPerInterval$meanSteps,type='l',
     main="Average Daily Activity Pattern", xlab="Interval", ylab="Mean steps")
```


```{r maxAveSteps, echo=TRUE}
# get the interval with the max number of steps
maxStepIndex <- which.max(averageStepsPerInterval$meanSteps);
maxInterval <- averageStepsPerInterval$interval[maxStepIndex]; 
```

The interval with the maximum number of steps per day is `r maxInterval`.





## Imputing missing values

```{r numberNA, echo=TRUE}
# get the interval with the max number of steps
numNaSteps <- sum((is.na(rawActivity$steps))==TRUE);
numNaInterval = sum((is.na(rawActivity$interval))==TRUE);
numNaDate = sum((is.na(rawActivity$date))==TRUE);

```

The number of NA step is  `r numNaSteps`.

The number of NA date is  `r numNaDate`.

The number of NA interval is  `r numNaInterval`.



To fill in the missing values the mean for that interval will be used.



```{r replaceMissing, echo=TRUE}

# find any intervals with NA steps and replace those values with the mean for that interval
# with more time, replace this loop with an sapply or similar command

# make a copy that we will fill
filledActivity <- rawActivity;

for(i in 1:length(rawActivity$steps))
  {
    if( is.na(rawActivity$steps[i]))
    {
       
      newValue <- averageStepsPerInterval$meanSteps[ (averageStepsPerInterval$interval == 
                                            rawActivity$interval[i])];
      
      filledActivity$steps[i]<-newValue;
    }
  
  }

# diagnostic - verify that all the values have been fixed
#count( is.na(filledActivity$steps) )


 
```


Now the basic statistics of the new data set will be checked.


```{r filledPlot, echo=TRUE}
# summarise the steps for each day and add back to the frame
sumFilledSteps<-ddply( filledActivity,c("date"),summarise, sumSteps=sum(steps))



# make a histogram of the step sums
hist(sumFilledSteps$sumSteps, main="Histogram of Filled Dataset Total Number of Steps Taken Each Day",
     xlab="Total Steps per Day");

```


```{r computeFilledMeanAndMedian, echo=TRUE}
# compute the mean and median for reporting in the main text
totalStepsMeanFilled <- mean(sumFilledSteps$sumSteps);
totalStepsMedianFilled <- median(sumFilledSteps$sumSteps);

# compute the delta from the unfilled set

deltaMean <- totalStepsMean - totalStepsMeanFilled;
deltaMedian <- totalStepsMedian - totalStepsMedianFilled;

```

The mean of the total number of steps per day is `r totalStepsMeanFilled`.

The median of the total number of steps per day is `r totalStepsMedianFilled`


 

The differnces between the filled and original set means of the total number of steps per day is `r deltaMean`.

The differnces between the filled and original set median of the total number of steps per day is `r deltaMedian`.


The impact on the mean is exactly zero as would be expected with a backfill using the mean from the intervals.   

The shift in the median is non-zero but vanishingly small given the total number of steps per day.   It is literally on order of a single missed step in a day.






 
## Are there differences in activity patterns between weekdays and weekends?


Generate a new version of the data set with a new column for "weekday" or "weekend".


```{r generateWeekdayColumn, echo=TRUE}

library(lattice)

# first generate the day of the week
filledActivity$day<- weekdays(as.Date(filledActivity$date));

# fill new column with default
filledActivity$weekday<-"weekday";

# now set either weekend or weekday
for(i in 1:length(filledActivity$day)){
  if(filledActivity$day[i] == "Sunday"){
    filledActivity$weekday[i] <- "weekend";
  }
  else if(filledActivity$day[i] == "Saturday"){
    filledActivity$weekday[i] <- "weekend";
  }
  # else leave it as the default weekday
}



# get the mean steps per interval
averageStepsPerWeekInterval<-ddply(filledActivity,c("weekday","interval"),summarise, meanSteps=mean(steps))



# do the plot
xyplot(   meanSteps ~ interval | factor(weekday)  ,
        data = averageStepsPerWeekInterval, type='l',
        xlab= "Interval", ylab = "Number of Steps");

 

 


```



